Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-08 multi_body_interactions/papers/theory/reviews_NJP/referee_1_response.tex

\renewcommand{\c}{\hat{c}}
\newcommand{\n}{\hat{n}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\K}{\mathcal{K}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\1}{\mathds{1}}

% inline lists
\usepackage[inline]{enumitem}
\setlist[enumerate]{leftmargin=*} % nice margins for enumerate

% for feynman diagrams
\usepackage{tikz,tikz-feynman}
\tikzset{
  baseline = (current bounding box.center)
}
\tikzfeynmanset{
  compat = 1.1.0,
  every feynman = {/tikzfeynman/small}
}
\newcommand{\shrink}[1]{\scalebox{0.8}{#1}} % for smaller diagrams

@@

\begin{document}

\section*{Referee 1}

We thank the referee for taking the time to carefully read our
manuscript, and for providing thoughtful and constructive feedback.
Below, we address the comments and concerns brought up by the referee.
We hope that the referee finds our responses and revisions
satisfactory, deeming our manuscript acceptable for publication in New
Journal of Physics.

Excerpts from the referee reports are written in \blue{blue}, excerpts
from our original manuscript are written in \red{red}, and new text in
the revised version of our manuscript is written in \green{green}.
All page, line, and reference numbers generally refer to those in our
original manuscript, unless explicitly stated otherwise.

@@

\begin{enumerate}
\item Concerning:

  \blue{The authors study the effective on-site interactions for the
    description of several fermionic particles confined in a very deep
    optical lattice. The focus is on alkaline earth atoms, where the
    s-wave interaction strength is independent on the nuclear spin,
    and on all atoms in the lowest state of the trapping
    potential. The manuscript is an extended version on the
    theoretical tools applied in the comparison between experiment and
    theory of Ref. [34].  The analysis is very important and
    timely. However, in the present form, the manuscript has some
    significant short comings, which have to be improved before a
    decision about publication can be made.}

  \blue{One of the main points of criticism is that the manuscript is
    very poorly written and very difficult to access. A few examples
    are listed below:}

  We thank the referee for the succinct summary of our work, and for
  recognizing that our analysis is ``very important and timely''.  We
  hope that thew revisions outlined below have made our manuscript
  more accessible, and that we have addressed shortcomings identified
  by the referee to a satisfactory degree.

@@

  \red{... nuclear spin levels.  We fully characterize...}

  now reads:

  \green{... nuclear spin levels.  Our theory is limited to the
    experimental regime of (i) a deep lattice, with (ii) at most one
    atom occupying each nuclear spin state on any lattice site, where
    the latter restriction is a consequence of initial ground-state
    preparation.  We fully characterize...}

  We then added explicit mention of ground-state preparation to the
  introduction, replacing (page 4, line 22):

@@

  \red{... through the exquisite capabilities with OLCs.}

  and now:

  \green{... through the exquisite capabilities with OLCs.  Our theory
    is limited to the experimental regime of at most one atom
    occupying each nuclear spin state on any lattice site, which is a
    consequence initial ground-state preparation.}

  In order to motivate a later restriction to one orbital excitation
  per lattice site, we have rearranged and rewritten some of the next
  paragraph in the introduction, changing the text (page 4, line 33):

@@

  which now reads:

  \green{Though effective multi-body interactions have previously been
    studied in the context of harmonically [35, 36] and
    lattice-confined [37] neutral bosons prepared in a single
    hyperfine state, our work deals for the first time with fermions
    that have internal degrees of freedom and multiple collisional
    parameters. We perform a direct comparison between the many-body
    interaction energies predicted by our low-energy effective theory
    and the experimental measurements of the density-dependent orbital
    excitation spectra performed in ref.~[34], similarly to the
    measurements with bosons performed in ref.~[38].  To facilitate
    this comparison of excitation spectra and characterize the
    low-lying excitations in our effective theory, we consider a
    restriction of the theory to states with at most one orbital
    excitation per lattice site, and find that the SU($N$) symmetry of
    atomic collisions allow effective multi-body interactions to take
    a remarkably simple form. We note that past work has also detected
    experimental signatures of multi-body interactions in the form of
    quantum phase revivals [39].}

  We also clarify the limitations of our theory in the overview
  between Eqs.~(8) and (9), as discussed later in point
  \ref{pt:confusion}, and have made our technical contributions more
  explicit in the preview of our main results (in section II),
  replacing (page 7, line 48):

  \red{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align*}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M, \tag{9}
    \end{align*}
    where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear
    spin of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum
    terminates at $2I+1$ because this is the largest number of atoms
    which may initially occupy a single lattice site.  Focusing on the
    low-lying orbital state excitations of this theory, we find that
    SU($N$) symmetry allows us to express multi-body Hamiltonians in
    the simple form}

  \green{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align*}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
      \tag{9}
    \end{align*}
    where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
    coupling constants $G_X$, $I$ is the total nuclear spin of each
    atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum terminates at
    $2I+1$ because this is the largest number of atoms which may
    initially occupy a single lattice site.  We explicitly compute all
    $M$-body Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through order
    $p=3$, yielding effective two-, three-, and four-body
    interactions.  To (i) facilitate a comparison with the
    experimental measurements of many-body orbital excitation spectra
    performed in ref.~[34] and (ii) characterize the low-lying
    excitations in our effective theory, we also consider a
    restriction of the multi-body Hamiltonians $H_M$ to states with at
    most one orbital excitation per lattice site.  We find that the
    SU($N$) symmetry of atomic collisions allows us to express
    multi-body Hamiltonians in the simple form}

  In section IV (page 18), we have similarly changed the first
  sentence from (page 18, line line 45):

@@

  \red{Current experiments with ultracold ${}^{87}$Sr on a lattice can
    coherently address ground states and their low-lying orbital
    excitations for up to five atoms per lattice site [34].  Due to
    the SU($N$) symmetry of inter-atomic interactions, manifest in the
    fact that all coupling constants are independent of nuclear spin,
    we can write all effective $M$-body Hamiltonians which address
    these states in the form}

  to:

  \green{Current experiments with ultracold ${}^{87}$Sr on a lattice
    can coherently address ground states and single orbital
    excitations of up to five atoms per lattice site [34].  Due to the
    SU($N$) symmetry of inter-atomic interactions, manifest in the
    fact that all coupling constants are independent of nuclear spin,
    a restriction of the $M$-body Hamiltonians $H_M=\sum_p H_M^{(p)}$
    to the subspace of experimentally addressed states takes the form}

  Finally, we have modified some text in our summary to emphasize that
  our main theory is general with respect to orbital excitations, and
  that the restriction to one orbital excitation per lattice site is
  made to facilitate comparison with experiments; specifically, we
  have changed (page 28, line 40):

@@

  \green{Consider now a collection of fermionic atoms with total
    nuclear spin $I$ and two atomic orbital states loaded into a
    translationally invariant 3-D optical lattice.  While an external
    trapping potential will generally break discrete translational
    symmetry of the lattice, any background inhomogeneity can be made
    negligible by spectroscopically addressing a sufficiently small
    region of the lattice [34].  Throughout this paper, we work
    strictly in the deep-lattice regime with negligible tunneling
    between lattice sites, and assume that both atomic orbital states
    experience identical lattice potentials.}

  To reflect this change to section II, we have modified the
  description of section II on page 4, line 52, previously:

@@

  \red{In section II we provide an overview of the relevant one- and
    two-body physics of ultracold atoms on a lattice, and summarize
    our main result of effective multi-body interactions between these
    atom.}

  and now:

  \green{In section II we summarize the experimental procedures
    relevant to our work, provide an overview of the one- and two-body
    physics of ultracold atoms in a deep lattice, and preview our main
    technical results.}

@@

  \label{pt:confusion}

  We thank the referee for their careful reading of our manuscript,
  and for pointing out that our discussion in the paragraph between
  Eqs.~(8) and (9) is poorly worded.  The paragraph in question
  appears on page 7, lines 20--47:

  \red{We are interested in the ground and low-lying orbital
    eigenstates of several interacting atoms on a single lattice site,
    as such states can be readily prepared and coherently addressed in
    current experiments with ultracold atoms [34].  Although these
    experiments operate at temperatures sufficiently low for
    negligible thermal occupation of excited motional states, simply
    neglecting excited single-particle motional states fails to
    reproduce the observed spectra of multiply-occupied lattice sites.
    The reason for this failure lies in the fact that the bare
    two-body Hamiltonian $H_{\t{int}}$ in (7) is not diagonal in the
    basis of single-particle motional states.  Even at zero
    temperature, therefore, the presence of excited motional states
    modifies many-body atomic energy spectra when the magnitude of
    off-diagonal elements of $H_{\t{int}}$ approach the scale of
    low-lying single-particle motional state energy gaps.  To address
    this problem, we develop a low-energy effective theory for
    ultracold two-level fermionic atoms on a lattice.  As our work is
    motivated by progress in experiments which initialize all atoms in
    identical (i.e.~ground) orbital and motional states, to simplify
    our theory we assume that any $N$ atoms on a single lattice site
    occupy $N$ nuclear spin states.  Multiple occupation of a single
    nuclear spin state on the same lattice site is initially forbidden
    by fermionic statistics, and subsequently cannot be violated in
    the absence of tunneling and hyperfine coupling.}

  We have re-written this paragraph in its entirety to walk through
  the arguments and explanations therein more carefully, replacing it
  with the following text:

@@

  \green{In order to understand interacting atoms' orbital excitation
    spectrum in the context of a zero-temperature Hubbard-type model
    that eliminates non-dynamical (i.e.~ground-state) motional degrees
    of freedom, we develop a low-energy effective theory of
    interacting AEAs in a deep lattice.  To simplify our theory, we
    assume that any $N$ atoms on a single lattice site occupy $N$
    nuclear spin states.  Ground-state preparation of the atoms in an
    experiment implies that multiple occupation of a single nuclear
    spin state on any given lattice site is initially forbidden by
    fermionic statistics, and subsequently cannot occur in the absence
    of inter-site effects or hyperfine coupling between nuclear spin
    states.}

  Accordingly, we have modified language in section III that makes
  reference to the re-written paragraph above, changing (page 9, line
  45):

@@

  \red{The naive idea of neglecting all excited states and using
    $H_{\t{int}}$ directly to describe the spectrum of single-particle
    motional ground states is thus equivalent to truncating our
    expansion for $H_{\t{int}}^{\t{eff}}$ at first order.}

  to:

  \green{Writing down a single-band Hubbard model that simply neglects
    excited atomic motional states and uses $H_{\t{int}}$ directly to
    describe the orbital spectrum of interacting atoms is thus
    equivalent to truncating our expansion for $H_{\t{int}}^{\t{eff}}$
    at first order.}

  We hope that the new text between Eqs.~(8) and (9) is more readable
  and easier to understand.

@@

\item Concerning:

  \blue{(iii) Eq. (10) and Eq. (42) are identical.}

  As the referee points out, Eqs.~(10) and (42) in the original
  version of our manuscript are identical.  In order to avoid
  redundancy while still providing a preview of our main results, we
  have replaced the following text in section II (page 8, line 10):

  \red{\begin{align*} H_M = \sum_{\abs{\set{\mu_j}}=M} \p{U_{M,\g}
        \prod_{\alpha=1}^M \n_{\mu_\alpha,\g} + U_{M,+} \n_{\mu_1,\e}
        \prod_{\alpha=2}^M \n_{\mu_\alpha,\g} + U_{M,-}
        \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag \c_{\mu_2,\g}
        \c_{\mu_1,\e} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g}},
      \tag{10}
    \end{align*}
    where $\c_{\mu s}\equiv\c_{0,\mu s}$ is a fermionic annihilation
    operator for an atom in the motional ground state with nuclear
    spin $\mu$ and orbital state $s$;
    $\n_{\mu s}\equiv \c_{\mu s}^\dag \c_{\mu s}$ is a number
    operator; and the sum is performed over all choices of nuclear
    spins $\mu_j$ with $j=1,2,\cdots,M$ for which all $\mu_j$ are
    distinct, or equivalently all choices of $\mu_j$ for which the set
    $\set{\mu_j}$ contains $M$ elements, for a total of
    ${2I+1\choose M}\times\p{M!}$ nuclear spin combinations.  The
    coefficients $U_X$ can be expressed in terms of the coupling
    constants of the bare two-body interaction Hamiltonian (see
    Appendix F).  The key feature of the $M$-body interactions in (10)
    is that they take a form identical to the two-body interactions
    between motional ground-state atoms, but with the addition of
    $M-2$ spectator atoms.}

@@

  by:

  \green{\begin{align*}
      H_M = \sum_{\abs{\set{\mu_j}}=M}
      H_2^{(\mu_1,\mu_2)} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g},
      \tag{10}
    \end{align*}
    where $H_2^{(\mu_1,\mu_2)}$ is a two-body Hamiltonian addressing
    atoms with nuclear spin $\mu_1,\mu_2$;
    $\n_{\mu s}=\c_{\mu s}^\dag\c_{\mu s}$ is a number operator for
    atoms with nuclear state $\mu$ and orbital state $s$; and the sum
    is performed over all choices of nuclear spins $\mu_j$ with
    $j=1,2,\cdots,M$ for which all $\mu_j$ are distinct, or
    equivalently all choices of $\mu_j$ for which the set
    $\set{\mu_j}$ contains $M$ elements, for a total of
    ${2I+1\choose M}\times\p{M!}$ nuclear spin combinations.  The key
    feature of the $M$-body interactions in (10) is that they
    ultimately take the same form as two-body interactions, but with
    the addition of $M-2$ spectator atoms.}

  and we have modified Eq.~(42) to take the explicit form advertised
  in the new version of Eq.~(10), changing (page 18, line 54):

  \red{\begin{align*}
      H_M = \sum_{\abs{\set{\mu_j}}=M}
      \p{U_{M,\g} \prod_{\alpha=1}^M \n_{\mu_\alpha,\g}
        + U_{M,+} \n_{\mu_1,\e} \prod_{\alpha=2}^M \n_{\mu_\alpha,\g}
        + U_{M,-} \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag
        \c_{\mu_2,\g} \c_{\mu_1,\e} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g}},
      \tag{42}
    \end{align*}}

@@

  to:

  \green{\begin{align*}
      H_M = \sum_{\abs{\set{\mu_j}}=M}
      \p{U_{M,\g} \n_{\mu_1,\g} \n_{\mu_2,\g}
        + U_{M,+} \n_{\mu_1,\e} \n_{\mu_2,\g}
        + U_{M,-} \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag
        \c_{\mu_2,\g} \c_{\mu_1,\e}}
      \prod_{\alpha=3}^M \n_{\mu_\alpha,\g},
      \tag{44}
    \end{align*}}

  Note that the new equation number is 44 due to the addition of two
  new equations in response to point \ref{pt:divergence} below.

@@

\item Concerning the use of an interaction pseudopotential:

  \blue{(iv) It is well established that Eq.(2) is incorrect as the
    s-wave interaction in three-dimensions is described by a
    pseudo-potential and not a delta-function interaction. It is
    rather misleading that the authors to not point out the problem at
    the relevant position, but only afterwards require a
    renormalization of the divergencies without explaining the reason
    for their appearance.}

  \label{pt:pseudo-potential}

  We thank the referee for their attention to detail, and for pointing
  out the potential to mislead readers with respect to the nature of
  $s$-wave interactions by our expression of Eq.~(2).  We have
  therefore added the a paragraph to section II, following the
  re-statement of Eq.~(2) in Eq.~(5).  Specifically, the following
  text has been inserted in between lines 35 and 36 of page 6:

  \green{Note that the Hamiltonian in (5) is not the true microscopic
    interaction Hamiltonian of AEAs, but rather a generic form for a
    low-energy effective field theoretic description of two-body
    interactions [17, 21, 35--37, 42--44].  There are therefore two
    important points to keep in mind concerning our use of (5) to
    describe two-body interactions.  First, the use of effective field
    theory generically gives rise to divergences that must be dealt
    with either through regularization, e.g.~of the zero-range
    interaction potential implicitly assumed in the expression of (5)
    [45], or through renormalization of the coupling constants in the
    theory.  We chose the latter approach, as we will in any case find
    it convenient to renormalize the coupling constants in the
    effective theory developed in section III.  The choice of method
    to regulate divergences has no effect on the underlying physics.}

### 2019-02-08 multi_body_interactions/papers/theory/paper.tex
finished first draft of response to referee 1 from NJP

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\K}{\mathcal{K}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\1}{\mathds{1}}

@@

The remainder of this paper is structured as follows.  In section
\ref{sec:overview} we summarize the experimental procedures relevant
to our work, provide an overview one- and two-body physics of
ultracold atoms on a lattice, and preview our main technical results.
In section \ref{sec:low_energy} we discuss our method for deriving a
low-energy effective theory, provide a perturbative expansion for the
net effective Hamiltonian, and compute all $M$-body Hamiltonians
through third order in the low-energy effective theory.  We then
analyze the low-lying excitations of the effective theory in section
\ref{sec:spectra}, comparing spectral predictions with experimental
measurements, and study the orbital-state dynamics of nuclear spin
mixtures interrogated via Rabi spectroscopy.  Finally, we summarize
and conclude our findings in section \ref{sec:summary}, and provide
some discussion of future outlooks.

@@

Note that the Hamiltonian in \eqref{eq:H_int_fields} is not the true
microscopic interaction Hamiltonian of AEAs, but rather a generic form
for a low-energy effective field theoretic description of two-body
interactions \cite{scazza2014observation, cazalilla2009ultracold,
  johnson2012effective, yin2014universal, johnson2009effective,
  gorshkov2010twoorbital, wall2013strongly, busch1998two}.  There are
therefore two important points to keep in mind concerning our use of
\eqref{eq:H_int_fields} to describe two-body interactions.  First, the
use of effective field theory generically gives rise to divergences
that must be dealt with either through regularization, e.g.~of the
zero-range interaction potential implicitly assumed in the expression
of \eqref{eq:H_int_fields} \cite{giorgini2008theory}, or through
renormalization of the coupling constants in the theory.  We chose the
latter approach, as we will in any case find it convenient to
renormalize the coupling constants in the effective theory developed
in section \ref{sec:low_energy}.  Second, \eqref{eq:H_int_fields} is
only the first term in a low-energy expansion of two-body interactions
in effective field theory, which generally includes additional terms
containing derivatives of field operators.  Derivative terms
correspond to the dependence of two-body scattering on the relative
momentum $k$ of particles involved, with $k\to0$ in the zero-energy
limit.  In the present case of $s$-wave scattering, the leading
dependence of the two-body interaction Hamiltonian on the relative
momentum $k$ can be captured by use of an energy-dependent
pseudo-potential, which amounts to using a $k$-dependent effective
scattering length \cite{blume2002fermi}.  This effective scattering
length can be determined by expanding the $s$-wave collisional phase
shift in powers of the relative momentum $k$ \cite{giorgini2008theory,
  flambaum1999analytical}.  Details of this expansion will depend on
the characteristic length scale of finite-range interactions.  In our
work, these corrections to \eqref{eq:H_int_fields} will be relevant
only for the calculation of two-body interaction energies, appearing
at third order in the coupling constants $G_X$.  As we are primarily
interested in $M$-body interactions for $M\ge3$, we defer this
calculation to Appendix \ref{sec:momentum_dependence}.  We note that
our approach of using an unregularized contact potential,
renormalizing coupling constants, and separately accounting for
momentum-dependent scattering is essentially the same as that used for
similar calculations in refs.~\cite{johnson2012effective,
  yin2014universal, johnson2009effective}.

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Low-energy effective theory}
\label{sec:low_energy}

The net Hamiltonian $H = H_0 + H_{\t{int}}$ for interacting AEAs on a
lattice is not diagonal with respect to single-particle motional state
indices.  The problem of determining interacting atoms' orbital
excitation spectrum therefore nominally involves all atomic motional
degrees of freedom.  At zero temperature, however, each orbital state
of interacting atoms is associated with a single motional ground
state.  In order to compute an orbital excitation spectrum at zero
temperature, in principle we need to identify these motional ground
states.  We can then ignore all excited motional states, which will be
neither thermally occupied nor externally interrogated.  Such a
procedure would drastically reduce the dimensionality of the Hilbert
space necessary to describe the atoms, thereby greatly simplifying any
description of the atoms' orbital spectrum and dynamics.  In practice,
however, identifying the motional ground states of interacting atoms
and writing down a Hamiltonian restricted to this subspace is a very
difficult process to carry out analytically.

@@

Substituting the definition of $\I$ into
\eqref{eq:H_int_1_2}-\eqref{eq:H_int_3} yields expressions that are
highly reminiscent of standard non-degenerate perturbation theory in
quantum mechanics, but which nonetheless exhibit crucial differences.
The first, and most obvious difference is that these expressions are
operator equations, and that the sums over virtual states are
performed over a basis for the orthogonal complement of the subspace
$\H_{\t{ground}}^{\t{single}}$, rather than a basis for the orthogonal
complement of a single state, as in non-degenerate perturbation
theory.  Second, the non-degeneracy condition in standard perturbation
theory is here elevated to a restriction on the magnitude of the
perturbation $H_{\t{int}}$ relative to the spectral gap $\Delta$ of
the non-interacting Hamiltonian $H_0$ between
$\H_{\t{ground}}^{\t{single}}$ and the excited subspace $\E_0$.
Specifically, the validity of \eqref{eq:H_int_1_2}-\eqref{eq:H_int_3}
is conditional only on $\norm{H_{\t{int}}}\le\Delta/2$, where
$\norm{X}\equiv\max_{\ket\psi\in\H}\sqrt{\bk{\psi|X^\dag X|\psi}}$ is
the operator norm, with no restrictions on spectral gaps or
degeneracies within $\H_{\t{ground}}^{\t{single}}$ \cite{davis1969new,
  bravyi2011schrieffer}.  Finally, the effective theory involves no
corrections to the non-interacting many-body energy eigenstates; the
purpose of the effective Hamiltonian
$H_{\t{eff}} = H_0 + H_{\t{int}}^{\t{eff}}$ is to reproduce, on the
eigenstates of the non-interacting Hamiltonian $H_0$ within
$\H_{\t{ground}}^{\t{single}}$, the spectrum of the interacting
Hamiltonian $H = H_0 + H_{\t{int}}$ on $\H_{\t{ground}}^{\t{multi}}$.
Modifying the eigenstates of the non-interacting Hamiltonian $H_0$ on
$\H_{\t{ground}}^{\t{single}}$ thus invalidates the effective theory.

@@

The form of the bare two-body interaction Hamiltonian $H_{\t{int}}$ in
\eqref{eq:H_int} motivates a diagrammatic representation of terms in
the effective Hamiltonian $H_{\t{eff}}$ in \eqref{eq:H_int_eff},
similarly to the diagrams used to represent elements of the scattering
matrix in field theory.  Effective $M$-body interaction terms at order
$p$ in $H_{\t{int}}$ can be represented by directed graphs containing
$p$ internal vertices.  Each internal vertex and its associated edges
correspond respectively to a coupling constant and the associated
field operators in $H_{\t{int}}$.  An example 2-vertex diagram
representing an effective 3-body interaction term is provided in
figure \ref{fig:diagram}.  All diagrams are read from left to right to
construct a sequence of operators from right to left; the internal
vertices of a diagram are thus strictly ordered, with the $n$-th
internal vertex from the left corresponding to the $n$-th interaction
Hamiltonian $H_{\t{int}}$ from the right in \eqref{eq:H_int_1_2} or
\eqref{eq:H_int_3}.  Solid (dashed) lines represent field operators
acting on the lowest (arbitrary) motional states.  Spatial overlap
factors at each vertex are determined by the motional states of the
edges which connect to (i.e.~field operators associated with) that
vertex.  While it is possible to construct explicit rules for
determining the energetic suppression factors (i.e.~from $\I$) of the
term represented by a diagram, these factors are most easily
determined by examination of the effective Hamiltonians in
\eqref{eq:H_int_1_2} and \eqref{eq:H_int_3}.

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\K}{\mathcal{K}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}

\newcommand{\1}{\mathds{1}}

@@

The remainder of this paper is structured as follows.  In section
\ref{sec:overview} we summarize the experimental procedures relevant
to our work, provide an overview of the one- and two-body physics of
ultracold atoms in a deep lattice, and preview our main technical
results.  In section \ref{sec:low_energy} we discuss our method for
deriving a low-energy effective theory, provide a perturbative
expansion for the net effective Hamiltonian, and compute all $M$-body
Hamiltonians through third order in the low-energy effective theory.
We then analyze the low-lying excitations of the effective theory in
section \ref{sec:spectra}, comparing spectral predictions with
experimental measurements, and study the orbital-state dynamics of
nuclear spin mixtures interrogated via Rabi spectroscopy.  Finally, we
summarize and conclude our findings in section \ref{sec:summary}, and
provide some discussion of future outlooks.

@@

Note that the Hamiltonian in \eqref{eq:H_int_fields} is not the true
microscopic interaction Hamiltonian of AEAs, but rather a generic form
for a low-energy effective field theoretic description of two-body
interactions \cite{scazza2014observation, cazalilla2009ultracold,
  johnson2012effective, yin2014universal, johnson2009effective,
  gorshkov2010twoorbital, wall2013strongly, busch1998two}.  There are
therefore two important points to keep in mind concerning our use of
\eqref{eq:H_int_fields} to describe two-body interactions.  First, the
use of effective field theory generically gives rise to divergences
that must be dealt with either through regularization, e.g.~of the
zero-range interaction potential implicitly assumed in the expression
of \eqref{eq:H_int_fields} \cite{giorgini2008theory}, or through
renormalization of the coupling constants in the theory.  We chose the
latter approach, as we will in any case find it convenient to
renormalize the coupling constants in the effective theory developed
in section \ref{sec:low_energy}.  The choice of method to regulate
divergences has no effect on the underlying physics.

Second, \eqref{eq:H_int_fields} is only the first term in a low-energy
expansion of two-body interactions in effective field theory, which
generally includes additional terms containing derivatives of field
operators.  Derivative terms correspond to the dependence of two-body
scattering on the relative momentum $k$ of particles involved, with
$k\to0$ in the zero-energy limit.  In the present case of $s$-wave
scattering, the leading dependence of the two-body interaction
Hamiltonian on the relative momentum $k$ can be captured by use of an
energy-dependent pseudo-potential, which amounts to using a
$k$-dependent effective scattering length \cite{blume2002fermi}.  This
effective scattering length can be determined by expanding the
$s$-wave collisional phase shift in powers of the relative momentum
$k$ \cite{giorgini2008theory, flambaum1999analytical}.  Details of
this expansion will depend on the characteristic length scale of
finite-range interactions.  In our work, these corrections to
\eqref{eq:H_int_fields} will be relevant only for the calculation of
two-body interaction energies, appearing at third order in the
coupling constants $G_X$.  As we are primarily interested in $M$-body
interactions for $M\ge3$, we defer this calculation to Appendix
\ref{sec:momentum_dependence}.  We note that our approach of using an
unregularized contact potential, renormalizing coupling constants, and
separately accounting for momentum-dependent scattering is essentially
the same as that used for similar calculations in
refs.~\cite{johnson2012effective, yin2014universal,
  johnson2009effective}.  While this approach does not provide insight
into the microscopic structure of inter-atomic interactions, it is
suitable for the phenomenological description of these interactions,
and particular for our eventual development of a low-energy effective
theory.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Low-energy effective theory}
\label{sec:low_energy}

The net Hamiltonian $H = H_0 + H_{\t{int}}$ for interacting AEAs on a
lattice is not diagonal with respect to single-particle motional state
indices (e.g.~$n\in\mathbb{N}_0^3$).  The problem of determining
interacting atoms' orbital excitation spectrum therefore nominally
involves all atomic motional degrees of freedom.  At zero temperature,
however, each orbital state of a collection of interacting atoms is
associated with a single motional ground state.  In order to compute
an orbital excitation spectrum at zero temperature, in principle we
need to identify these motional ground states.  We can then ignore all
excited motional states, which will be neither thermally occupied nor
externally interrogated.  Such a procedure would drastically reduce
the dimensionality of the Hilbert space necessary to describe the
atoms, thereby greatly simplifying any description of the atoms'
orbital spectrum and internal (i.e.~nuclear and orbital) dynamics.  In
practice, however, identifying the motional ground states of
interacting atoms and writing down a Hamiltonian restricted to this
subspace is a very difficult process to carry out analytically.

@@

Substituting the definition of $\I$ into
\eqref{eq:H_int_1_2}-\eqref{eq:H_int_3} yields expressions that are
highly reminiscent of standard non-degenerate perturbation theory in
quantum mechanics, but which nonetheless exhibit crucial differences.
The first, and most obvious difference is that these expressions are
operator equations, and that the sums over virtual states are
performed over a basis for the orthogonal complement of the subspace
$\H_{\t{ground}}^{\t{single}}$, rather than a basis for the orthogonal
complement of a single state, as in non-degenerate perturbation
theory.  Second, the non-degeneracy condition in standard perturbation
theory is here elevated to a restriction on the magnitude of the
perturbation $H_{\t{int}}$ relative to the spectral gap $\Delta$ of
the non-interacting Hamiltonian $H_0$ between
$\H_{\t{ground}}^{\t{single}}$ and the excited subspace $\E_0$.
Specifically, the validity of \eqref{eq:H_int_1_2}-\eqref{eq:H_int_3}
is conditional only on $\norm{H_{\t{int}}}\le\Delta/2$, where
$\norm{X}\equiv\max_{\ket\psi\in\H}\sqrt{\bk{\psi|X^\dag X|\psi}}$ is
the operator norm, with no restrictions on spectral gaps or
degeneracies within $\H_{\t{ground}}^{\t{single}}$ \cite{davis1969new,
  bravyi2011schrieffer}.  Finally, the effective theory involves no
corrections to the non-interacting many-body energy eigenstates; the
purpose of constructing the effective Hamiltonian
$H_{\t{eff}} = H_0 + H_{\t{int}}^{\t{eff}}$ is to reproduce, on the
known eigenstates of the non-interacting Hamiltonian $H_0$ within
$\H_{\t{ground}}^{\t{single}}$, the spectrum of the interacting
Hamiltonian $H = H_0 + H_{\t{int}}$ on $\H_{\t{ground}}^{\t{multi}}$.
``Correcting'' the eigenstates of the non-interacting Hamiltonian
$H_0$ on $\H_{\t{ground}}^{\t{single}}$ thus invalidates the effective
theory.

@@

The form of the bare two-body interaction Hamiltonian $H_{\t{int}}$ in
\eqref{eq:H_int} motivates a diagrammatic representation of terms in
the effective Hamiltonian $H_{\t{eff}}$ in \eqref{eq:H_int_eff},
similarly to the diagrams used to represent elements of the scattering
matrix in standard quantum field theory.  Effective $M$-body
interaction terms at order $p$ in $H_{\t{int}}$ can be represented by
directed graphs containing $p$ internal vertices.  Each internal
vertex and its associated edges correspond respectively to a coupling
constant and the associated field operators in $H_{\t{int}}$.  An
example 2-vertex diagram representing an effective 3-body interaction
term is provided in figure \ref{fig:diagram}.  All diagrams are read
from left to right to construct a sequence of operators from right to
left; the internal vertices of a diagram are thus strictly ordered,
with the $n$-th internal vertex from the left corresponding to the
$n$-th interaction Hamiltonian $H_{\t{int}}$ from the right in
\eqref{eq:H_int_1_2} or \eqref{eq:H_int_3}.  Solid (dashed) lines
represent field operators acting on the lowest (arbitrary) motional
states.  Spatial overlap factors at each vertex are determined by the
motional states of the edges which connect to (i.e.~field operators
associated with) that vertex.  While it is possible to construct
explicit rules for determining the energetic suppression factors
(i.e.~from $\I$) of the term represented by a diagram, these factors
are most easily determined by examination of the effective
Hamiltonians in \eqref{eq:H_int_1_2} and \eqref{eq:H_int_3}.
